%!TEX root = ../template.tex

\chapter{License \& Acknowledgments}
\label{chap:license}

\section{License Overview}

The \gls{novathesis} Template, its associated files, and this User Manual are distributed under the terms of the \textbf{LaTeX Project Public License (LPPL)}, version~1.3c or later.  
The LPPL is an established license recognized by the Free Software Foundation, designed specifically for the distribution and maintenance of \LaTeX{} classes, packages, and documentation.

This license ensures that the template remains freely available for academic and research use, while preserving author attribution and version integrity.

\paragraph{Key Provisions of the LPPL.}
\begin{itemize}
  \item Users are free to use, copy, and distribute the template and documentation;
  \item Modifications are permitted, provided that:
    \begin{itemize}
      \item The modified file is clearly identified as such;
      \item The original author’s name and copyright notice are retained;
      \item The modified file is not distributed under the same name as the original;
    \end{itemize}
  \item The software is provided “as is,” without any express or implied warranty;
  \item The LPPL encourages active maintenance and transparency in academic projects.
\end{itemize}

\paragraph{License Reference.}
A copy of the full license text is included in the repository as \texttt{LICENSE.txt} and can also be found at:

\begin{quote}
\url{https://www.latex-project.org/lppl.txt}
\end{quote}

\paragraph{License Declaration.}
\begin{quote}
\textcopyright{}
2008–2025, NOVA University Lisbon and individual contributors.  
This work may be distributed and/or modified under the conditions of the  
\textbf{LaTeX Project Public License (LPPL)}, version~1.3c or later.
\end{quote}

\section{Authorship and Maintenance}

The \gls{novathesis} project was conceived, developed, and is maintained by the academic and technical community of \textbf{NOVA University Lisbon}.  
Its long-term sustainability is ensured by continuous institutional support and voluntary contributions from students, faculty, and researchers.

\paragraph{Core Maintainers.}
\begin{itemize}
  \item \gls{novathesis} Coordination Team — NOVA University Lisbon
  \item Faculty of Science and Technology (FCT NOVA)
  \item Collaborating Schools and Research Units
\end{itemize}

\paragraph{Editorial Oversight.}
\begin{itemize}
  \item Documentation maintained by the \gls{novathesis} Documentation Committee;
  \item Peer review provided by representatives of participating faculties.
\end{itemize}

Contributors are listed individually in the \texttt{AUTHORS.md} file included with the repository.

\section{Acknowledgment of Contributions}

\gls{novathesis} incorporates the work and feedback of numerous contributors who assisted in the design, testing, and translation of the template.  
The maintainers extend their gratitude to all individuals who provided issue reports, feature suggestions, and localization improvements over the years.

\paragraph{Community Contributions.}
\begin{itemize}
  \item Developers of institutional presets and cover layouts;
  \item Translators responsible for multilingual support;
  \item Testers validating compatibility across TeX~Live, MacTeX, and MiKTeX;
  \item Early adopters and graduate students who provided field feedback;
  \item Academic supervisors and reviewers who ensured compliance with official guidelines.
\end{itemize}

\paragraph{Upstream Software.}
\gls{novathesis} depends on the foundational work of the following \LaTeX{} packages and projects:
\begin{itemize}
  \item \texttt{memoir} — document class for books and theses;
  \item \texttt{biblatex} and \texttt{biber} — bibliography and citation management;
  \item \texttt{fontspec} and \texttt{unicode-math} — modern font and math support;
  \item \texttt{hyperref} and \texttt{cleveref} — hyperlinking and cross-referencing;
  \item \texttt{glossaries}, \texttt{siunitx}, and \texttt{tikz} — advanced typesetting and figures.
\end{itemize}

The maintainers acknowledge the entire \LaTeX{} development community for their continued contributions to the open-source academic publishing ecosystem.

\section{Institutional Support}

\gls{novathesis} was originally developed to standardize thesis formatting at NOVA University Lisbon and has since evolved into a reference model for academic writing across multiple institutions.

Institutional support is provided by:
\begin{itemize}
  \item NOVA University Lisbon Rectorate;
  \item Faculty of Science and Technology (FCT NOVA);
  \item NOVA School of Social Sciences and Humanities (FCSH NOVA);
  \item NOVA Information Management School (NOVA IMS);
  \item NOVA School of Business and Economics (NOVA SBE);
  \item Associated research centers and graduate programs.
\end{itemize}

Their continued engagement ensures that the template reflects current formatting, linguistic, and submission standards.

\section{Citation of the \gls{novathesis} Template}

When referencing the \gls{novathesis} Template or this manual in publications, please use the following citation format:

\begin{quote}
NOVA University Lisbon.  
\emph{\gls{novathesis} Template and User Manual}.  
Version~7.6.0, April~2025.  
Available at \url{https://github.com/novathesis/novathesis}.
\end{quote}

This citation ensures proper academic attribution and version traceability.

\section{Disclaimer}

The \gls{novathesis} Template and its documentation are provided for academic use without warranty of any kind.  
The maintainers and affiliated institutions are not responsible for any damages or non-compliance resulting from improper configuration, modification, or institutional requirements that deviate from the template defaults.

Users must verify that their final thesis complies with the official formatting guidelines of their specific department or graduate school.

\section{Final Note}

\gls{novathesis} exemplifies the values of open academic collaboration—free distribution, transparent governance, and respect for institutional diversity.  
Its continued success depends on contributions from the academic community and adherence to open-source principles.

\bigskip
\noindent\textit{Lisbon, November~2025}\\
\textit{\gls{novathesis} Coordination Team}\\
\textit{NOVA University Lisbon}
